hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper.Rmd
edits to overview and ef framework

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for their creation arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

$$
Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
$$

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum of effective values reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

>>>>

**hatchR** is designed primarily as a tool for predicting fish phenology. To maintain focus on this core function, we provide minimal build-in data validation and vizualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensize validation checks. However, we include basic data-checking functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`). Intuitive functions are provided for users to apply models (either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function). Users can then apply these models to water temperature data (e.g. from a HOBO temp logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external vizualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows–particulalry for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.

@@

Effective value models were introduced by @sparks2019 to predict developmental timing in wild populations, initially for Sockeye Salmon (*O. nerka*). Their development was necessitated by limitations in traditional models, such as those in @beacham1990, which relied on average incubation temperature over the full developmental period. In wild settings, estimating this average temperature was impracticable since hatch timing was unknown, even when spawning dates were recorded. To overcome this challenge, @sparks2019 reformulated model 2 from @beacham1990 by taking its reciprocal and assigning an *effective value* to each day of development based on the daily average temperature. This approach allowed for cumulative tracking of developmental progress, enabling hatch and emergence predictions without requiring prior knowledge of incubation temperatures.

@@

$$
\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b)}
$$

where $E_i$ is the effective value and $T_i$ the temperature for day $i$, and $a$ and $b$ are model parameterization estimates (i.e. species- or model-specific constants).

This framework as the foundation for phenological models in **hatchR**. The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979). While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function. This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.
